# DATA SOURCE:
# Connecticut Crash Data Repository
# https://www.ctcrash.uconn.edu/

# GEOGRAPHIC SCOPE:
# Hartford County / Greater Hartford-area crashes

# TIMEFRAME:
# January 1 to August 28, 2026

# CRASH DATASETS:
# Dataset 1 = Crash-level data
# Dataset 2 = Vehicle-level data
# Dataset 3 = Person/driver-level data

# ANALYSIS BASE:
# Drivers from Dataset 3

# JOIN KEYS:
# Dataset 1 -> Dataset 3: CrashId
# Dataset 2 -> Dataset 3: CrashId + VehicleId

# PROBLEM STATEMENT:
# Which environmental, roadway, and vehicle factors are associated
# with both driver injury and vehicle damage severity
# in Greater Hartford-area crashes during 2026?

# DEPENDENT VARIABLE 1:
# Driver_Injury
# Source: Dataset 3
# 0 = No apparent injury
# 1 = Injury or fatal injury

# DEPENDENT VARIABLE 2:
# Vehicle_Damage_Severity
# Source: Dataset 2
# Based on Extent of Damage Text Format

# Candidate independent variables will be finalized
# after data fusion, missing-value checks,
# correlation analysis, and the 0.50 cutoff.

library(ggplot2)
library(ggplotAssist)
library(rmarkdown)
library(data.table)

getwd()

#assigning and importing datasets

crash01 <- "https://raw.githubusercontent.com/TasbiHasan/Project-Crashdata/refs/heads/main/export_215822_0.csv"
crash02 <- "https://raw.githubusercontent.com/TasbiHasan/Project-Crashdata/refs/heads/main/export_215822_1.csv"
crash03 <- "https://github.com/TasbiHasan/Project-Crashdata/raw/refs/heads/main/export_215822_2.csv"

#putting in dataframe
#I tried with read.table but that didn't help. Tried to debug and got "fread" as suggestion

ctable01 <- fread(file = crash01, header = TRUE, sep = ",")
ctable02 <- fread(file = crash02, header = TRUE, sep = ",")
ctable03 <- fread(file = crash03, header = TRUE, sep = ",")


```{r}
# Create working copies of the raw datasets

crash_data <- copy(ctable01)
vehicle_data <- copy(ctable02)
person_data <- copy(ctable03)
```


```{r}

#getting rid of column duplication dataset 01-dataset 02 

names(crash_data) <- make.unique(
  names(crash_data),
  sep = "_duplicate"
)

names(vehicle_data) <- make.unique(
  names(vehicle_data),
  sep = "_duplicate"
)

names(person_data) <- make.unique(
  names(person_data),
  sep = "_duplicate"
)

anyDuplicated(names(crash_data))
anyDuplicated(names(vehicle_data))
anyDuplicated(names(person_data))
```

```{r}

#Narrowing it down to only drivers


table(
  person_data$`Person Type Text Format`,
  useNA = "ifany"
)

driver_data <- person_data[
  `Person Type Text Format` == "Driver"
]


nrow(driver_data)

uniqueN(driver_data$CrashId)

head(
  driver_data[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`
  )]
)
```

#Driver Identification

```{r}
anyDuplicated(
  driver_data[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)
```
#if we can fuse the data with crash_data
```{r}
all(
  driver_data$CrashId %in%
    crash_data$CrashId
)
```

```{r}
#creating temporary matching IDs


driver_vehicle_key <- paste(
  driver_data$CrashId,
  driver_data$VehicleId,
  sep = "_"
)

vehicle_key <- paste(
  vehicle_data$CrashId,
  vehicle_data$VehicleId,
  sep = "_"
)


all(
  driver_vehicle_key %in%
    vehicle_key
)
```

#checking the columns we need before fusing to Dataset -3

```{r}
# Columns we want from Dataset 1

needed_crash_cols <- c(
  "CrashId",
  "Date Of Crash",
  "Time of Crash",
  "Latitude",
  "Longitude",
  "Hour of the Day",
  "Light Condition Text Format",
  "Road Surface Condition Text Format",
  "Type of Intersection Text Format"
)

needed_crash_cols %in% names(crash_data)

setdiff(
  needed_crash_cols,
  names(crash_data)
)
```


```{r}
crash_selected <- crash_data[, ..needed_crash_cols]

nrow(crash_selected)

uniqueN(crash_selected$CrashId)

anyDuplicated(crash_selected$CrashId)
```

```{r}
driver_crash <- merge(
  driver_data,
  crash_selected,
  by = "CrashId",
  all.x = TRUE,
  sort = FALSE
)

nrow(driver_crash)

uniqueN(driver_crash$CrashId)

anyDuplicated(
  driver_crash[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)



sum(is.na(driver_crash$`Date Of Crash`))

sum(is.na(driver_crash$Latitude))

sum(is.na(driver_crash$Longitude))




head(
  driver_crash[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`,
    `Date Of Crash`,
    `Time of Crash`,
    Latitude,
    Longitude,
    `Light Condition Text Format`,
    `Road Surface Condition Text Format`,
    `Type of Intersection Text Format`
  )]
)
```




```{r}
#looking into dataset 2

needed_vehicle_cols <- c(
  "CrashId",
  "VehicleId",
  "Posted/Statutory Speed Limit",
  "Vehicle Model Year",
  "Extent of Damage Text Format",
  "Total Lanes In Roadway",
  "Roadway Alignment Text Format",
  "Roadway Grade Text Format"
)

needed_vehicle_cols %in% names(vehicle_data)

setdiff(
  needed_vehicle_cols,
  names(vehicle_data)
)
```

```{r}
#looking for variables 

table(
  vehicle_data$`Extent of Damage Text Format`,
  useNA = "ifany"
)


table(
  vehicle_data$`Posted/Statutory Speed Limit`,
  useNA = "ifany"
)

summary(
  vehicle_data$`Vehicle Model Year`
)

table(
  vehicle_data$`Total Lanes In Roadway`,
  useNA = "ifany"
)


```



```{r}
vehicle_selected <- vehicle_data[, ..needed_vehicle_cols]




#one vehicle appears once

anyDuplicated(
  vehicle_selected[, .(
    CrashId,
    VehicleId
  )]
)

```


```{r}
#cleaning speed limit variable


valid_speed_limits <- c(
  5, 10, 15, 20, 25, 30, 35,
  40, 45, 50, 55, 60, 65
)

vehicle_selected[, Speed_Limit_Clean :=
  fifelse(
    `Posted/Statutory Speed Limit` %in% valid_speed_limits,
    as.numeric(`Posted/Statutory Speed Limit`),
    NA_real_
  )
]


table(
  vehicle_selected$Speed_Limit_Clean,
  useNA = "ifany"
)
```



```{r}
#the vehicle-damage dependent variable


vehicle_selected[, Damage_Score :=
  fcase(
    `Extent of Damage Text Format` == "No Damage", 0,
    `Extent of Damage Text Format` == "Minor Damage", 1,
    `Extent of Damage Text Format` == "Functional Damage", 2,
    `Extent of Damage Text Format` == "Disabling Damage", 3,
    default = NA_real_
  )
]

table(
  vehicle_selected$Damage_Score,
  useNA = "ifany"
)

```

#checking model year of vehicles


```{r}
table(
  vehicle_selected[
    `Vehicle Model Year` < 1980 |
    `Vehicle Model Year` > 2026,
    `Vehicle Model Year`
  ],
  useNA = "ifany"
)

vehicle_selected[
  `Vehicle Model Year` < 1980 |
  `Vehicle Model Year` > 2026,
  .N,
  by = `Vehicle Model Year`
][order(`Vehicle Model Year`)]
```

#checking the different model years

```{r}
vehicle_data[
  `Vehicle Model Year` < 1980 |
  `Vehicle Model Year` > 2026,
  .(
    CrashId,
    VehicleId,
    Make,
    Model,
    `Vehicle Model Year`,
    `Vehicle Unit Type Text Format`,
    `Body Type Text Format`
  )
][order(`Vehicle Model Year`)]



sum(is.na(vehicle_selected$`Vehicle Model Year`))
```

```{r}
#Model year Distribution

quantile(
  vehicle_selected$`Vehicle Model Year`,
  probs = c(0, .01, .05, .25, .50, .75, .95, .99, 1),
  na.rm = TRUE
)
```


#cleaning unusual vehicle model year


```{r}
vehicle_selected[, Vehicle_Model_Year_Clean :=
  fifelse(
    `Vehicle Model Year` == 1900,
    NA_real_,
    as.numeric(`Vehicle Model Year`)
  )
]

table(
  vehicle_selected$Vehicle_Model_Year_Clean,
  useNA = "ifany"
)

sum(is.na(vehicle_selected$Vehicle_Model_Year_Clean))
```

#variable Vehicle Age


```{r}
vehicle_selected[, Vehicle_Age :=
  2026 - Vehicle_Model_Year_Clean
]

vehicle_selected[
  Vehicle_Age < 0,
  Vehicle_Age := 0
]


summary(vehicle_selected$Vehicle_Age)

quantile(
  vehicle_selected$Vehicle_Age,
  probs = c(0, .01, .05, .25, .50, .75, .95, .99, 1),
  na.rm = TRUE
)
```
#merging dataset 2 into driver_dataset


```{r}
driver_vehicle_crash <- merge(
  driver_crash,
  vehicle_selected,
  by = c("CrashId", "VehicleId"),
  all.x = TRUE,
  sort = FALSE
)



#damage checking

nrow(driver_vehicle_crash)

uniqueN(driver_vehicle_crash$CrashId)

anyDuplicated(
  driver_vehicle_crash[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)
```
```{r}
sum(is.na(driver_vehicle_crash$Damage_Score))

sum(is.na(driver_vehicle_crash$Speed_Limit_Clean))

sum(is.na(driver_vehicle_crash$Vehicle_Age))

```

```{r}
head(
  driver_vehicle_crash[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`,
    Damage_Score,
    Speed_Limit_Clean,
    Vehicle_Age,
    `Road Surface Condition Text Format`,
    `Light Condition Text Format`
  )]
)


```

```{r}
nrow(driver_vehicle_crash)

uniqueN(driver_vehicle_crash$CrashId)

anyDuplicated(
  driver_vehicle_crash[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)

sum(is.na(driver_vehicle_crash$Damage_Score))
sum(is.na(driver_vehicle_crash$Speed_Limit_Clean))
sum(is.na(driver_vehicle_crash$Vehicle_Age))
```

```{r}
#Importing dataset

# Weather data file paths

weather01 <- "C:/TASBI PROJECT ANA/weather data-20260922T095903Z-1-001/weather data/GHCNh_USW00014740_2026_converted.csv"

weather02 <- "C:/TASBI PROJECT ANA/weather data-20260922T095903Z-1-001/weather data/GHCNh_USW00014752_2026.csv"

```




```{r}

#Putting it into dataframe 
weather1 <- fread(
  file = weather01,
  header = TRUE,
  sep = ","
)

weather2 <- fread(
  file = weather02,
  header = TRUE,
  sep = ","
)

```


```{r}
nrow(weather1)
ncol(weather1)

nrow(weather2)
ncol(weather2)
```


```{r}
head(names(weather1), 20)

head(names(weather2), 20)
```



```{r}
#checking if both dataset has the same column

identical(
  names(weather1),
  names(weather2)
)


identical(
  names(weather1),
  names(weather2)
)
```

#checking weather stations 

```{r}
unique(
  weather1[, .(
    STATION,
    Station_name,
    LATITUDE,
    LONGITUDE
  )]
)

unique(
  weather2[, .(
    STATION,
    Station_name,
    LATITUDE,
    LONGITUDE
  )]
)
```





#we need to inspect the date and time before we go further


```{r}
class(weather1$DATE)
head(weather1$DATE)
tail(weather1$DATE)
```

```{r}
class(weather2$DATE)
head(weather2$DATE)
tail(weather2$DATE)
```

#Imoprtant weather variable for us

```{r}
summary(weather1$temperature)
summary(weather1$wind_speed)
summary(weather1$visibility)

summary(weather2$temperature)
summary(weather2$wind_speed)
summary(weather2$visibility)
```


```{r}

#checking how much is missing 

sum(is.na(weather1$temperature))
sum(is.na(weather1$wind_speed))
sum(is.na(weather1$visibility))

sum(is.na(weather2$temperature))
sum(is.na(weather2$wind_speed))
sum(is.na(weather2$visibility))
```
#Combine the two weather datasets


```{r}
weather_all <- rbindlist(
  list(weather1, weather2),
  use.names = TRUE
)

nrow(weather_all)
ncol(weather_all)

uniqueN(weather_all$STATION)
```


#Create a smaller weather table

```{r}
weather_core <- weather_all[, .(
  STATION,
  Station_name,
  DATE,
  LATITUDE,
  LONGITUDE,
  temperature,
  wind_speed,
  visibility
)]


nrow(weather_core)

head(weather_core)
```


#Checking how often the three useful weather variables appear together


```{r}
weather_core[, .(
  Temperature_Available = sum(!is.na(temperature)),
  Wind_Available = sum(!is.na(wind_speed)),
  Visibility_Available = sum(!is.na(visibility))
)]

sum(
  complete.cases(
    weather_core[, .(
      temperature,
      wind_speed,
      visibility
    )]
  )
)
```

```{r}
sum(
  !is.na(weather_core$temperature) |
  !is.na(weather_core$wind_speed) |
  !is.na(weather_core$visibility)
)
```

#Checking for duplicate station timestamps


```{r}
weather_duplicates <- weather_core[
  ,
  .N,
  by = .(STATION, DATE)
][N > 1]


nrow(weather_duplicates)

head(weather_duplicates)
```
#crash date and time format 


```{r}
class(driver_vehicle_crash$`Date Of Crash`)
class(driver_vehicle_crash$`Time of Crash`)


head(driver_vehicle_crash$`Date Of Crash`)
head(driver_vehicle_crash$`Time of Crash`)
```

#creating the crash data and time 

```{r}
driver_vehicle_crash[, Crash_DateTime :=
  as.POSIXct(
    paste(`Date Of Crash`, `Time of Crash`),
    format = "%Y-%m-%d %H:%M:%S",
    tz = "America/New_York"
  )
]
```

```{r}
head(driver_vehicle_crash$Crash_DateTime)

class(driver_vehicle_crash$Crash_DateTime)

sum(is.na(driver_vehicle_crash$Crash_DateTime))
```
```{r}
attr(weather_core$DATE, "tzone")

head(weather_core$DATE)
```


#comparing the timezone


```{r}
head(
  format(
    driver_vehicle_crash$Crash_DateTime,
    tz = "America/New_York",
    usetz = TRUE
  )
)


head(
  format(
    driver_vehicle_crash$Crash_DateTime,
    tz = "UTC",
    usetz = TRUE
  )
)
```


#creating data table


```{r}
weather_match <- weather_core[
  !is.na(temperature) &
  !is.na(wind_speed) &
  !is.na(visibility)
]
```


```{r}
nrow(weather_match)

uniqueN(weather_match$STATION)

range(weather_match$DATE)
```

```{r}
weather_match[, .N, by = STATION]
```

```{r}
unique(
  weather_match[, .(
    STATION,
    Station_name,
    LATITUDE,
    LONGITUDE
  )]
)
```

#Create a crash-level weather matching table
```{r}
crashes_for_weather <- unique(
  driver_vehicle_crash[, .(
    CrashId,
    Crash_DateTime,
    Crash_Latitude = Latitude,
    Crash_Longitude = Longitude
  )]
)
```


```{r}
nrow(crashes_for_weather)

uniqueN(crashes_for_weather$CrashId)

anyDuplicated(crashes_for_weather$CrashId)
```

#Create a table containing the two weather stations

```{r}
weather_stations <- unique(
  weather_match[, .(
    STATION,
    Station_name,
    Station_Latitude = LATITUDE,
    Station_Longitude = LONGITUDE
  )]
)

print(weather_stations)
nrow(weather_stations)
```

#now creating distance function


```{r}
haversine_km <- function(lat1, lon1, lat2, lon2) {
  
  R <- 6371
  
  lat1 <- lat1 * pi / 180
  lon1 <- lon1 * pi / 180
  lat2 <- lat2 * pi / 180
  lon2 <- lon2 * pi / 180
  
  dlat <- lat2 - lat1
  dlon <- lon2 - lon1
  
  a <- sin(dlat / 2)^2 +
    cos(lat1) * cos(lat2) *
    sin(dlon / 2)^2
  
  c <- 2 * atan2(
    sqrt(a),
    sqrt(1 - a)
  )
  
  R * c
}
```

#now we would calculate the distance between both stations

#first station 

```{r}
crashes_for_weather[, Distance_Station1_km :=
  haversine_km(
    Crash_Latitude,
    Crash_Longitude,
    weather_stations$Station_Latitude[1],
    weather_stations$Station_Longitude[1]
  )
]
```

#second station 

```{r}
crashes_for_weather[, Distance_Station2_km :=
  haversine_km(
    Crash_Latitude,
    Crash_Longitude,
    weather_stations$Station_Latitude[2],
    weather_stations$Station_Longitude[2]
  )
]
```

```{r}
summary(
  crashes_for_weather$Distance_Station1_km
)

summary(
  crashes_for_weather$Distance_Station2_km
)
```

#now we will assign the nearest station 

```{r}
crashes_for_weather[, Nearest_Station :=
  fifelse(
    Distance_Station1_km <= Distance_Station2_km,
    weather_stations$STATION[1],
    weather_stations$STATION[2]
  )
]
```

#and creating the actual nearest distance

```{r}
crashes_for_weather[, Nearest_Station_Distance_km :=
  pmin(
    Distance_Station1_km,
    Distance_Station2_km
  )
]
```


```{r}
table(
  crashes_for_weather$Nearest_Station
)

summary(
  crashes_for_weather$Nearest_Station_Distance_km
)


quantile(
  crashes_for_weather$Nearest_Station_Distance_km,
  probs = c(0, .25, .50, .75, .90, .95, 1),
  na.rm = TRUE
)


```


#now we move to matching the time 


```{r}
weather_for_join <- weather_match[, .(
  STATION,
  Join_Time = DATE,
  Weather_Observed_UTC = DATE,
  temperature,
  wind_speed,
  visibility
)]

setkey(
  weather_for_join,
  STATION,
  Join_Time
)
```

```{r}
nrow(weather_for_join)

head(weather_for_join)
```



#preparing crash time for matching 


```{r}
crash_weather_request <- crashes_for_weather[, .(
  CrashId,
  STATION = Nearest_Station,
  Crash_DateTime,
  Crash_Time_UTC = as.POSIXct(
    as.numeric(Crash_DateTime),
    origin = "1970-01-01",
    tz = "UTC"
  ),
  Nearest_Station_Distance_km
)]

crash_weather_request[, Join_Time := Crash_Time_UTC]
```

```{r}
head(
  crash_weather_request[, .(
    CrashId,
    STATION,
    Crash_DateTime,
    Crash_Time_UTC,
    Nearest_Station_Distance_km
  )]
)
```



#matching each crash to nearest weather observation 


```{r}
weather_nearest <- weather_for_join[
  crash_weather_request,
  on = .(
    STATION,
    Join_Time
  ),
  roll = "nearest"
]



nrow(weather_nearest)

uniqueN(weather_nearest$CrashId)

anyDuplicated(weather_nearest$CrashId)
```

#Calculating how far apart the crash and weather times are


```{r}
weather_nearest[, Weather_Time_Diff_Min :=
  abs(
    as.numeric(
      difftime(
        Weather_Observed_UTC,
        Crash_Time_UTC,
        units = "mins"
      )
    )
  )
]

summary(
  weather_nearest$Weather_Time_Diff_Min
)




quantile(
  weather_nearest$Weather_Time_Diff_Min,
  probs = c(
    0,
    .25,
    .50,
    .75,
    .90,
    .95,
    .99,
    1
  ),
  na.rm = TRUE
)
```

```{r}
weather_nearest[, .(
  Within_30_Min = sum(Weather_Time_Diff_Min <= 30),
  Within_60_Min = sum(Weather_Time_Diff_Min <= 60),
  Within_90_Min = sum(Weather_Time_Diff_Min <= 90),
  Over_90_Min   = sum(Weather_Time_Diff_Min > 90)
)]
```

#worst matches 

```{r}
head(
  weather_nearest[
    order(-Weather_Time_Diff_Min),
    .(
      CrashId,
      STATION,
      Crash_DateTime,
      Weather_Observed_UTC,
      Weather_Time_Diff_Min
    )
  ],
  10
)
```


#checking how many are over 60 mins time gap


```{r}
weather_nearest[, .(
  Within_30_Min = sum(Weather_Time_Diff_Min <= 30),
  Within_60_Min = sum(Weather_Time_Diff_Min <= 60),
  Over_60_Min   = sum(Weather_Time_Diff_Min > 60)
)]
```
```{r}

weather_nearest[
  Weather_Time_Diff_Min > 60,
  .(
    CrashId,
    STATION,
    Crash_DateTime,
    Weather_Observed_UTC,
    Weather_Time_Diff_Min,
    temperature,
    wind_speed,
    visibility
  )
][order(-Weather_Time_Diff_Min)]

```



#We should not delete the crashes.Instead, if the nearest weather reading is more than 60 minutes away, we mark the weather values as missing


```{r}
weather_nearest[
  Weather_Time_Diff_Min > 60,
  `:=`(
    temperature = NA_real_,
    wind_speed = NA_real_,
    visibility = NA_real_
  )
]
```
#We should not pretend that old observation represents the crash weather.



```{r}
sum(is.na(weather_nearest$temperature))
sum(is.na(weather_nearest$wind_speed))
sum(is.na(weather_nearest$visibility))
```

#creating final weather columns 

```{r}
weather_final <- weather_nearest[, .(
  CrashId,
  Weather_Station = STATION,
  Weather_Station_Distance_km = Nearest_Station_Distance_km,
  Weather_Time_Diff_Min,
  Temperature = temperature,
  Wind_Speed = wind_speed,
  Visibility = visibility
)]
```


```{r}
nrow(weather_final)

uniqueN(weather_final$CrashId)

anyDuplicated(weather_final$CrashId)
```



#FUSING THE DATASET


```{r}
final_fused_data <- merge(
  driver_vehicle_crash,
  weather_final,
  by = "CrashId",
  all.x = TRUE,
  sort = FALSE
)

```

```{r}
nrow(final_fused_data)

uniqueN(final_fused_data$CrashId)

anyDuplicated(
  final_fused_data[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)
```
#checking the weather dataset

```{r}
sum(is.na(final_fused_data$Temperature))
sum(is.na(final_fused_data$Wind_Speed))
sum(is.na(final_fused_data$Visibility))



summary(final_fused_data$Temperature)
summary(final_fused_data$Wind_Speed)
summary(final_fused_data$Visibility)





```

```{r}

#checking driver injury categories 

table(
  final_fused_data$`Injury Status Text Format`,
  useNA = "ifany"
)
```

#creating DEPENDENT VARIABLE 01 

```{r}

final_fused_data[, Driver_Injury :=
  fcase(
    `Injury Status Text Format` == "No Apparent Injury (O)", 0,
    
    `Injury Status Text Format` %in% c(
      "Possible Injury (C)",
      "Suspected Minor Injury (B)",
      "Suspected Serious Injury (A)",
      "Fatal Injury (K)"
    ), 1,
    
    default = NA_real_
  )
]

#review
table(
  final_fused_data$Driver_Injury,
  useNA = "ifany"
)


#percentage

prop.table(
  table(final_fused_data$Driver_Injury)
) * 100
```

#checking both targets togather

```{r}
table(
  Driver_Injury = final_fused_data$Driver_Injury,
  Damage_Score = final_fused_data$Damage_Score,
  useNA = "ifany"
)
```

#We would inspect our variables to check if we can use them 


```{r}
table(
  final_fused_data$`Light Condition Text Format`,
  useNA = "ifany"
)

table(
  final_fused_data$`Road Surface Condition Text Format`,
  useNA = "ifany"
)

table(
  final_fused_data$`Type of Intersection Text Format`,
  useNA = "ifany"
)
```

#numericals

```{r}
table(
  final_fused_data$`Roadway Alignment Text Format`,
  useNA = "ifany"
)

table(
  final_fused_data$`Roadway Grade Text Format`,
  useNA = "ifany"
)
```



#numericals 


```{r}
summary(final_fused_data$Speed_Limit_Clean)

summary(final_fused_data$Vehicle_Age)

summary(final_fused_data$`Total Lanes In Roadway`)

summary(final_fused_data$Temperature)

summary(final_fused_data$Wind_Speed)

summary(final_fused_data$Visibility)

summary(final_fused_data$`Hour of the Day`)
```


#checking missing values 

```{r}
final_fused_data[, .(
  Missing_Speed_Limit = sum(is.na(Speed_Limit_Clean)),
  Missing_Vehicle_Age = sum(is.na(Vehicle_Age)),
  Missing_Total_Lanes = sum(is.na(`Total Lanes In Roadway`)),
  Missing_Temperature = sum(is.na(Temperature)),
  Missing_Wind = sum(is.na(Wind_Speed)),
  Missing_Visibility = sum(is.na(Visibility))
)]
```





#creating variable dark condition 

```{r}
final_fused_data[, Dark_Condition :=
  fcase(
    `Light Condition Text Format` %in% c(
      "Dark-Lighted",
      "Dark-Not Lighted",
      "Dark-Unknown Lighting"
    ), 1,

    `Light Condition Text Format` %in% c(
      "Daylight",
      "Dawn",
      "Dusk"
    ), 0,

    default = NA_real_
  )
]


table(
  final_fused_data$Dark_Condition,
  useNA = "ifany"
)
```


#Adverse Road Condition



```{r}
final_fused_data[, Adverse_Road :=
  fcase(
    `Road Surface Condition Text Format` == "Dry", 0,

    `Road Surface Condition Text Format` %in% c(
      "Wet",
      "Ice / Frost",
      "Slush",
      "Snow",
      "Standing Water",
      "Moving Water",
      "Mud, Dirt, Gravel",
      "Sand"
    ), 1,

    default = NA_real_
  )
]




table(
  final_fused_data$Adverse_Road,
  useNA = "ifany"
)

```


#Intersection indicator


```{r}

final_fused_data[, At_Intersection :=
  fcase(
    `Type of Intersection Text Format` ==
      "Not at Intersection", 0,

    `Type of Intersection Text Format` %in% c(
      "Four-Way Intersection",
      "T-Intersection",
      "Y-Intersection",
      "L-Intersection",
      "Five-Point, or More",
      "Roundabout",
      "Traffic Circle"
    ), 1,

    default = NA_real_
  )
]


table(
  final_fused_data$At_Intersection,
  useNA = "ifany"
)


```


#roadway alignment variable


```{r}
final_fused_data[, Curved_Road :=
  fcase(
    `Roadway Alignment Text Format` == "Straight", 0,

    `Roadway Alignment Text Format` %in% c(
      "Curve Left",
      "Curve Right"
    ), 1,

    default = NA_real_
  )
]


table(
  final_fused_data$Curved_Road,
  useNA = "ifany"
)
```


#roadway grade variable


```{r}


final_fused_data[, Non_Level_Road :=
  fcase(
    `Roadway Grade Text Format` == "Level", 0,

    `Roadway Grade Text Format` %in% c(
      "Downhill",
      "Uphill",
      "Hillcrest",
      "Sag (bottom)"
    ), 1,

    default = NA_real_
  )
]



table(
  final_fused_data$Non_Level_Road,
  useNA = "ifany"
)

```

```{r}
table(final_fused_data$Dark_Condition, useNA = "ifany")

table(final_fused_data$Adverse_Road, useNA = "ifany")

table(final_fused_data$At_Intersection, useNA = "ifany")

table(final_fused_data$Curved_Road, useNA = "ifany")

table(final_fused_data$Non_Level_Road, useNA = "ifany")
```


#cleaning lanes 


```{r}
final_fused_data[, Total_Lanes_Clean :=
  fifelse(
    `Total Lanes In Roadway` > 0,
    as.numeric(`Total Lanes In Roadway`),
    NA_real_
  )
]

table(
  final_fused_data$Total_Lanes_Clean,
  useNA = "ifany"
)
```
```{r}
summary(final_fused_data$Total_Lanes_Clean)
```

#building the first predictor dataset

```{r}
correlation_candidates <- final_fused_data[, .(
  Temperature,
  Wind_Speed,
  Visibility,
  Dark_Condition,
  Adverse_Road,
  At_Intersection,
  Speed_Limit_Clean,
  Vehicle_Age,
  Curved_Road,
  Non_Level_Road,
  Total_Lanes_Clean,
  Hour_of_Day = `Hour of the Day`
)]
```

#checking missing values before correlation matrix 

```{r}
colSums(
  is.na(correlation_candidates)
)
```

```{r}
nrow(correlation_candidates)

sum(
  complete.cases(correlation_candidates)
)
```

#correlation matrix 


```{r}
cor_matrix <- cor(
  correlation_candidates,
  use = "pairwise.complete.obs",
  method = "pearson"
)

round(cor_matrix, 2)
```

#finding the correlation above .50

```{r}
high_correlations <- which(
  abs(cor_matrix) >= 0.50 &
  abs(cor_matrix) < 1,
  arr.ind = TRUE
)
high_correlations

cor_matrix
```


#supecious lane 
```{r}
table(
  final_fused_data$Total_Lanes_Clean,
  useNA = "ifany"
)
```


```{r}
final_fused_data[
  Total_Lanes_Clean >= 10,
  .(
    CrashId,
    VehicleId,
    Total_Lanes_Clean,
    `Roadway Alignment Text Format`,
    Speed_Limit_Clean
  )
]
```


#creating the reduced 10 variable predictor set


```{r}
final_predictor_candidates <- final_fused_data[, .(
  Temperature,
  Wind_Speed,
  Dark_Condition,
  Adverse_Road,
  At_Intersection,
  Speed_Limit_Clean,
  Vehicle_Age,
  Curved_Road,
  Non_Level_Road,
  Total_Lanes_Clean
)]
```


#Running Correlation matrix again 



```{r}
final_cor_matrix <- cor(
  final_predictor_candidates,
  use = "pairwise.complete.obs",
  method = "pearson"
)

round(final_cor_matrix, 2)
```

#checking for high correlation 


```{r}
final_high_correlations <- which(
  abs(final_cor_matrix) >= 0.50 &
  abs(final_cor_matrix) < 1,
  arr.ind = TRUE
)

final_high_correlations
```

```{r}
colSums(
  is.na(final_predictor_candidates)
)

sum(
  complete.cases(final_predictor_candidates)
)
```

creating VIF dataset
```{r}
vif_data <- na.omit(
  final_predictor_candidates
)

nrow(vif_data)
```
```{r}
calculate_vif <- function(data) {
  
  vif_results <- numeric(ncol(data))
  names(vif_results) <- names(data)
  
  for (variable in names(data)) {
    
    other_variables <- setdiff(
      names(data),
      variable
    )
    
    formula_vif <- as.formula(
      paste(
        variable,
        "~",
        paste(other_variables, collapse = " + ")
      )
    )
    
    model_vif <- lm(
      formula_vif,
      data = data
    )
    
    r_squared <- summary(model_vif)$r.squared
    
    vif_results[variable] <- 1 / (1 - r_squared)
  }
  
  return(vif_results)
}


vif_results <- calculate_vif(vif_data)

round(vif_results, 2)



```



#DOWNLOADING THE DATA


```{r}
getwd()

fwrite(
  final_fused_data,
  "final_fused_driver_crash_weather_2026.csv"
)

file.exists(
  "final_fused_driver_crash_weather_2026.csv"
)

```

