# DATA SOURCE:
# Connecticut Crash Data Repository
# https://www.ctcrash.uconn.edu/

# GEOGRAPHIC SCOPE:
# Hartford County / Greater Hartford-area crashes

# TIMEFRAME:
# January 1 to August 28, 2026

# CRASH DATASETS:
# Dataset 1 = Crash-level data
# Dataset 2 = Vehicle-level data
# Dataset 3 = Person/driver-level data

# ANALYSIS BASE:
# Drivers from Dataset 3

# JOIN KEYS:
# Dataset 1 -> Dataset 3: CrashId
# Dataset 2 -> Dataset 3: CrashId + VehicleId

# PROBLEM STATEMENT:
# Which environmental, roadway, and vehicle factors are associated
# with both driver injury and vehicle damage severity
# in Greater Hartford-area crashes during 2026?

# DEPENDENT VARIABLE 1:
# Driver_Injury
# Source: Dataset 3
# 0 = No apparent injury
# 1 = Injury or fatal injury

# DEPENDENT VARIABLE 2:
# Vehicle_Damage_Severity
# Source: Dataset 2
# Based on Extent of Damage Text Format

# Candidate independent variables will be finalized
# after data fusion, missing-value checks,
# correlation analysis, and the 0.50 cutoff.

library(ggplot2)
library(ggplotAssist)
library(rmarkdown)
library(data.table)

getwd()

#assigning and importing datasets

crash01 <- "https://raw.githubusercontent.com/TasbiHasan/Project-Crashdata/refs/heads/main/export_215822_0.csv"
crash02 <- "https://raw.githubusercontent.com/TasbiHasan/Project-Crashdata/refs/heads/main/export_215822_1.csv"
crash03 <- "https://github.com/TasbiHasan/Project-Crashdata/raw/refs/heads/main/export_215822_2.csv"

#putting in dataframe
#I tried with read.table but that didn't help. Tried to debug and got "fread" as suggestion

ctable01 <- fread(file = crash01, header = TRUE, sep = ",")
ctable02 <- fread(file = crash02, header = TRUE, sep = ",")
ctable03 <- fread(file = crash03, header = TRUE, sep = ",")


```{r}
# Create working copies of the raw datasets

crash_data <- copy(ctable01)
vehicle_data <- copy(ctable02)
person_data <- copy(ctable03)
```


```{r}

#getting rid of column duplication dataset 01-dataset 02 

names(crash_data) <- make.unique(
  names(crash_data),
  sep = "_duplicate"
)

names(vehicle_data) <- make.unique(
  names(vehicle_data),
  sep = "_duplicate"
)

names(person_data) <- make.unique(
  names(person_data),
  sep = "_duplicate"
)

anyDuplicated(names(crash_data))
anyDuplicated(names(vehicle_data))
anyDuplicated(names(person_data))
```

```{r}

#Narrowing it down to only drivers


table(
  person_data$`Person Type Text Format`,
  useNA = "ifany"
)

driver_data <- person_data[
  `Person Type Text Format` == "Driver"
]


nrow(driver_data)

uniqueN(driver_data$CrashId)

head(
  driver_data[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`
  )]
)
```

#Driver Identification

```{r}
anyDuplicated(
  driver_data[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)
```
#if we can fuse the data with crash_data
```{r}
all(
  driver_data$CrashId %in%
    crash_data$CrashId
)
```

```{r}
#creating temporary matching IDs


driver_vehicle_key <- paste(
  driver_data$CrashId,
  driver_data$VehicleId,
  sep = "_"
)

vehicle_key <- paste(
  vehicle_data$CrashId,
  vehicle_data$VehicleId,
  sep = "_"
)


all(
  driver_vehicle_key %in%
    vehicle_key
)
```

#checking the columns we need before fusing to Dataset -3

```{r}
# Columns we want from Dataset 1

needed_crash_cols <- c(
  "CrashId",
  "Date Of Crash",
  "Time of Crash",
  "Latitude",
  "Longitude",
  "Hour of the Day",
  "Light Condition Text Format",
  "Road Surface Condition Text Format",
  "Type of Intersection Text Format"
)

needed_crash_cols %in% names(crash_data)

setdiff(
  needed_crash_cols,
  names(crash_data)
)
```


```{r}
crash_selected <- crash_data[, ..needed_crash_cols]

nrow(crash_selected)

uniqueN(crash_selected$CrashId)

anyDuplicated(crash_selected$CrashId)
```

```{r}
driver_crash <- merge(
  driver_data,
  crash_selected,
  by = "CrashId",
  all.x = TRUE,
  sort = FALSE
)

nrow(driver_crash)

uniqueN(driver_crash$CrashId)

anyDuplicated(
  driver_crash[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)



sum(is.na(driver_crash$`Date Of Crash`))

sum(is.na(driver_crash$Latitude))

sum(is.na(driver_crash$Longitude))




head(
  driver_crash[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`,
    `Date Of Crash`,
    `Time of Crash`,
    Latitude,
    Longitude,
    `Light Condition Text Format`,
    `Road Surface Condition Text Format`,
    `Type of Intersection Text Format`
  )]
)
```




```{r}
#looking into dataset 2

needed_vehicle_cols <- c(
  "CrashId",
  "VehicleId",
  "Posted/Statutory Speed Limit",
  "Vehicle Model Year",
  "Extent of Damage Text Format",
  "Total Lanes In Roadway",
  "Roadway Alignment Text Format",
  "Roadway Grade Text Format"
)

needed_vehicle_cols %in% names(vehicle_data)

setdiff(
  needed_vehicle_cols,
  names(vehicle_data)
)
```

```{r}
#looking for variables 

table(
  vehicle_data$`Extent of Damage Text Format`,
  useNA = "ifany"
)


table(
  vehicle_data$`Posted/Statutory Speed Limit`,
  useNA = "ifany"
)

summary(
  vehicle_data$`Vehicle Model Year`
)

table(
  vehicle_data$`Total Lanes In Roadway`,
  useNA = "ifany"
)


```



```{r}
vehicle_selected <- vehicle_data[, ..needed_vehicle_cols]




#one vehicle appears once

anyDuplicated(
  vehicle_selected[, .(
    CrashId,
    VehicleId
  )]
)

```


```{r}
#cleaning speed limit variable


valid_speed_limits <- c(
  5, 10, 15, 20, 25, 30, 35,
  40, 45, 50, 55, 60, 65
)

vehicle_selected[, Speed_Limit_Clean :=
  fifelse(
    `Posted/Statutory Speed Limit` %in% valid_speed_limits,
    as.numeric(`Posted/Statutory Speed Limit`),
    NA_real_
  )
]


table(
  vehicle_selected$Speed_Limit_Clean,
  useNA = "ifany"
)
```



```{r}
#the vehicle-damage dependent variable


vehicle_selected[, Damage_Score :=
  fcase(
    `Extent of Damage Text Format` == "No Damage", 0,
    `Extent of Damage Text Format` == "Minor Damage", 1,
    `Extent of Damage Text Format` == "Functional Damage", 2,
    `Extent of Damage Text Format` == "Disabling Damage", 3,
    default = NA_real_
  )
]

table(
  vehicle_selected$Damage_Score,
  useNA = "ifany"
)

```

#checking model year of vehicles


```{r}
table(
  vehicle_selected[
    `Vehicle Model Year` < 1980 |
    `Vehicle Model Year` > 2026,
    `Vehicle Model Year`
  ],
  useNA = "ifany"
)

vehicle_selected[
  `Vehicle Model Year` < 1980 |
  `Vehicle Model Year` > 2026,
  .N,
  by = `Vehicle Model Year`
][order(`Vehicle Model Year`)]
```

#checking the different model years

```{r}
vehicle_data[
  `Vehicle Model Year` < 1980 |
  `Vehicle Model Year` > 2026,
  .(
    CrashId,
    VehicleId,
    Make,
    Model,
    `Vehicle Model Year`,
    `Vehicle Unit Type Text Format`,
    `Body Type Text Format`
  )
][order(`Vehicle Model Year`)]



sum(is.na(vehicle_selected$`Vehicle Model Year`))
```

```{r}
#Model year Distribution

quantile(
  vehicle_selected$`Vehicle Model Year`,
  probs = c(0, .01, .05, .25, .50, .75, .95, .99, 1),
  na.rm = TRUE
)
```


#cleaning unusual vehicle model year


```{r}
vehicle_selected[, Vehicle_Model_Year_Clean :=
  fifelse(
    `Vehicle Model Year` == 1900,
    NA_real_,
    as.numeric(`Vehicle Model Year`)
  )
]

table(
  vehicle_selected$Vehicle_Model_Year_Clean,
  useNA = "ifany"
)

sum(is.na(vehicle_selected$Vehicle_Model_Year_Clean))
```

#variable Vehicle Age


```{r}
vehicle_selected[, Vehicle_Age :=
  2026 - Vehicle_Model_Year_Clean
]

vehicle_selected[
  Vehicle_Age < 0,
  Vehicle_Age := 0
]


summary(vehicle_selected$Vehicle_Age)

quantile(
  vehicle_selected$Vehicle_Age,
  probs = c(0, .01, .05, .25, .50, .75, .95, .99, 1),
  na.rm = TRUE
)
```
#merging dataset 2 into driver_dataset


```{r}
driver_vehicle_crash <- merge(
  driver_crash,
  vehicle_selected,
  by = c("CrashId", "VehicleId"),
  all.x = TRUE,
  sort = FALSE
)



#damage checking

nrow(driver_vehicle_crash)

uniqueN(driver_vehicle_crash$CrashId)

anyDuplicated(
  driver_vehicle_crash[, .(
    CrashId,
    VehicleId,
    PersonId
  )]
)
```
```{r}
sum(is.na(driver_vehicle_crash$Damage_Score))

sum(is.na(driver_vehicle_crash$Speed_Limit_Clean))

sum(is.na(driver_vehicle_crash$Vehicle_Age))

```

```{r}
head(
  driver_vehicle_crash[, .(
    CrashId,
    VehicleId,
    PersonId,
    Age,
    `Injury Status Text Format`,
    Damage_Score,
    Speed_Limit_Clean,
    Vehicle_Age,
    `Road Surface Condition Text Format`,
    `Light Condition Text Format`
  )]
)


```


